### 2009-06-01 response-to-reviews2.txt

Thank you for the detailed recommendations.  We now state clearly that
we take the phylogeny as given, although we also note in the
Discussion that in principle we could construct a basic sampler
over trees using our method (though we have not constructed or
evaluated such a sampler).

We have added more discussion of the potential rate-limiting steps and
corresponding constraints to the manuscript ("Overview").  As the
reviewer points out, perhaps the most realistic pieces of outside
information to have are a consensus structure for sequences and/or a
curated multiple alignment.  In principle these can all be
incorporated into our inference algorithms with the alignment
constraints used by, e.g., QRNA and PFOLD, or the simultaneous fold and
alignment constraints used by, e.g., Stemloc and CONSAN.

For this paper we used constraints on individual structures because
these are simple to implement.  We state this clearly in the revised
text and note that while experimentally-determined structures of
individual RNAs are rare, curated multiple alignments are
more-commonly available and can be used to constrain the algorithms
which we describe.

In principle we think that the reconstruction might benefit as well
from partial alignment constraints as from partial fold constraints;
that is, as well as being able to deviate slightly from the consensus
fold (as the reviewer suggests), the reconstruction algorithm would
probably need to deviate slightly from the supplied alignment as
well. (This will be necessary because not every evolutionarily related
basepair is likely to be perfectly aligned.) This sort of flexible
fold and alignment constraint is implemented in both CONSAN and
Stemloc. However, we note that the Pair SCFGs used in both those
algorithms were first implemented using much simpler constraints:
specifically, the fold-only constraints of (Holmes 2002), and the
alignment-only constraints of (Rivas & Eddy 2001).

We hope & anticipate that the sophistication of the constraint schemes
used with these sorts of model will steadily increase over
time. Eventually we hope to develop alignment constraints,
simultaneous alignment-and-fold constraints, and other constraints
(and/or sampling schemes) to enable efficient analysis of multiple
phylogenetically-related sequences. The present work represents a
first attempt at using these sorts of models for more than two
sequences, but we hope it will not be the last attempt, and we
acknowledge room for improvement.

@@

> It is still not clear eenough what the authors' model is, though it is 
> supposed to be a main result. Page 7 states "a prescriptive formalism 
> for transducers is given in Text S1". Relegating the deliverables of a 
> manuscript to supplementary material is not an appropriate use of 
> supplementary material; a reader should not have to refer to a second 
> manuscript to understand the main points. The authors need to find a 
> way to explain their main results in the actual manuscript. There is 
> still much vagueness and redundancy in this draft, which could be 
> removed and replaced with more relevant explanation. For example, in 
> this same section where results are relegated to supplementary 
> material, there's a whole redundant paragraph (p.8) on how conditional 
> models represent directed edges; but this concept has already been 
> explained sufficiently. This is followed (bottom p. 8, "Consider a 
> phylogenetic tree...") by yet another statement of the same concept of 
> using conditional models, still in
> terms too vague and conceptual to enable reproduction of the work.

Thank you for pointing this out.  We have attempted to make the
main text a self-contained description of the composition algorithm,
including its key corollaries (elimination of null states and DP
inference).  The presentation of the composition algorithm in the
supplement now merely reprises the presentation in the main text using
more formal notation. We hope this is an appropriate use of the
supplementary information, but we welcome additional suggestions to
improve the flow of the manuscript for the reader.

@@

We agree that these reconstructions were not very informative.
Unfortunately these were the only sequences that we could find which
were short enough to be tractable on the hardware we had available. We
now report simulation results instead, as suggested.

We thank the reviewer for pointing out errors in the original
figures. These were due to mistakes made during manual conversion of
indiegram's output parse trees into structural alignments and a
bug in the code used for maximum-likelihood reconstruction of
sequences (the xrate program). We have now automated this process and
fixed the bug, verifying that this has eliminated the errors
(in fact this verification is incorporated into our new benchmark).

### 2009-06-01 paper.tex
Edits per changes to supplementary material.

A notational prescription for the allowed transitions may be found in \supptext.

@@

\supptext\ includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
along with procedures for probabilistically restoring null cycles
to sampled parse trees and Inside-Outside expectation counts.
Informally, the essence of the algorithm is contained within the following two steps:
\begin{description}
\item[(i)] separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children;
\item[(ii)] replacing all empty paths through null states with effective direct transitions between non-null states,
obtaining sum-over-paths probabilities by inverting the grammar's transition matrix.
\end{description}

@@

@@

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Rule $\rho'' \in \grules''$}
  \KwOut{Parse (sub)tree $T' \in \gtrees'$}

  \Switch{$\rho''$}{
    \Case{$X \to Y$}{
      \eIf{\rnd $< P(X \to Y) / q_{XY}$}{
        \KwRet{$(X \to Y)$} \;
      }{
        Let $z = \sum_W q_{XW} P'(W \to Y)$ \;
        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to Y) / z$ \;
        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to Y)$} \;
      }
    }
    \Case{$X \to \epsilon$}{
      \eIf{\rnd $< P'(X \to \epsilon) / P'(\epsilon | X)$}{
        \KwRet{$(X \to \epsilon)$} \;
      }{
        Let $z = \sum_W q_{XW} P'(W \to \epsilon)$ \;
        Sample state $V$ from probability distribution $P(V) = q_{XV} P'(V \to \epsilon) / z$ \;
        \KwRet{$(\mathrm{restoreTransitions}(X \to V) \to \epsilon)$} \;
      }
    }
    \Other{
      \KwRet{$(\rho'')$} \;
    }
  }
  \caption{\label{alg:restoreTransitions}
    Subroutine $\mathrm{restoreTransitions}(\rho'')$ for the null cycle elimination procedure.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Rule $\rho' \in \grules'$}
  \KwOut{Parse (sub)tree $T \in \gtrees$}

  \Switch{$\rho'$}{
    \Case{$B^0 \to \epsilon$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
      \KwRet{$(B \to T_L\ T_R)$} \;
    }
    \Case{$B' \to L'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_R \leftarrow \mathrm{sampleNullSubtree}(R)$ \;
      \KwRet{$(B \to L\ T_R)$} \;
    }
    \Case{$B' \to R'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      $T_L \leftarrow \mathrm{sampleNullSubtree}(L)$ \;
      \KwRet{$(B \to T_L\ R)$} \;
    }
    \Case{$B^2 \to L'\ R'$}{
      Let $B \to L\ R$ be the original bifurcation in $\grules$ \;
      \KwRet{$(B \to L\ R)$} \;
    }
    \lCase{$B^0 \to B'$}{
      \KwRet{$()$} \;
    }
    \lCase{$B' \to B^2$}{
      \KwRet{$()$} \;
    }
    \Other{
      Let $\rho$ be the original rule in $\grules$ \;
      \KwRet{$(\rho)$} \;
    }
  }

  \caption{\label{alg:restoreBifurcations}
    Subroutine $\mathrm{restoreBifurcations}(\rho')$ for the null cycle elimination procedure.
  }
\end{algorithm}

\begin{algorithm}[!ht]
  \SetLine
  \SetKwFunction{rnd}{Random[0,1]}
  \KwIn{Nonterminal $A \in \gnons$}
  \KwOut{Parse tree $T \in \gtrees: \mathrm{seq}(T) = \epsilon, \mathrm{root}(T) = A$}

  \eIf{$A$ is a bifurcation state, $A \to X\ Y$,}{
    $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
    $T_Y \leftarrow \mathrm{sampleNullSubtree}(Y)$ \;
    \KwRet{$(A \rightarrow T_X\ T_Y)$} \;
  }{
    \eIf{\rnd $< P(A \to \epsilon) / P(\epsilon|A)$}{
      \KwRet{$(A \to \epsilon)$} \;
    }{
      Let $z = \sum_Y P(A \to Y) P(\epsilon | Y)$ \;
      Sample state $X$ from probability distribution $P(X) = P(A \to X) P(\epsilon | X) / z$ \;
      $T_X \leftarrow \mathrm{sampleNullSubtree}(X)$ \;
      \KwRet{$(A \rightarrow T_X)$} \;
    }
  }

  \caption{\label{alg:sampleNullSubtree}
    Subroutine $\mathrm{sampleNullSubtree}(A)$ for the null cycle elimination procedure.
  }
\end{algorithm}

>>>>

A notational prescription for the allowed transitions may be found in \supptext{1}.

@@

\supptext{2} includes a complete formal algorithm for exact null-cycle elimination in SCFGs,
along with procedures for probabilistically restoring null cycles
to sampled parse trees and Inside-Outside expectation counts.
Informally, the essence of the algorithm is contained within the following two steps:
\begin{description}
\item[(i)] separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children;
\item[(ii)] replacing all empty paths through null states with effective direct transitions between non-null states,
obtaining sum-over-paths probabilities by inverting the grammar's transition matrix.
\end{description}

@@

The available software is described in \supptext{3}.

@@
